---
title: "STAT115 Homework 1"
author: ""
date: "January 28, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)# please knit with `echo=TRUE, eval=TRUE`
```


# Part 0: Odyssey Signup

Please fill out the Odyssey survey on Canvas so we can create an account for you. 

# Part I: Introduction to R

## Problem 1: Installation

**Please install the following R/Bioconductor packages**

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install()
BiocManager::install("sva")

install.packages(c("ggplot2", "dplyr", "tidyr", "HistData", "mvtnorm",
                   "reticulate"))
```


Please run this command to see if Bioconductor can work fine.

```{r} 
# please knit with this command echoed.
BiocManager::valid() 
```


```{r libraries, message = FALSE}
# these packages are needed for HW2
# affy and affyPLM are needed to read the microarray data and run RMA
library(sva) # for batch effect correction. Contains ComBat and sva.
library(ggplot2) # for plotting
library(dplyr) # for data manipulation
library(reticulate) # needed to run python in Rstudio
# these next two are not essential to this course
library(mvtnorm) # need this to simulate data from multivariate normal
library(HistData) # need this for data
```

When we grade, we should download their Rmd and knit it ourselves. They
should have points removed if their document does not knit. On this HW,
we will not take off points, but will warn them for next time.

If they have hard-coded file names, we will warn them on this as well.

## Problem 2: Getting help

You can use the `mean()` function to compute the mean of a vector like
so:

```{r mean}
x1 <- c(1:10, 50)
mean(x1)
```

However, this does not work if the vector contains NAs:

```{r mean-na}
x1_na <- c(1:10, 50, NA)
mean(x1_na)
```

**Please use R documentation to find the mean after excluding NA's (hint: `?mean`)**

```{r problem2}
# your code here
```


# Part II: Data Manipulation

## Problem 3: Basic Selection

In this question, we will practice data manipulation using a dataset
collected by Francis Galton in 1886 on the heights of parents and their
children. This is a very famous dataset, and Galton used it to come up
with regression and correlation.

The data is available as `GaltonFamilies` in the `HistData` package.
Here, we load the data and show the first few rows. To find out more
information about the dataset, use `?GaltonFamilies`.

```{r loadGalton}
data(GaltonFamilies)
head(GaltonFamilies)
```

a. **Please report the height of the 10th child in the dataset.**

```{r problem3a}
# your code here
```

b. **What is the breakdown of male and female children in the dataset?**

```{r problem3b}
# your code here
```

c. **How many observations are in Galton's dataset? Please answer this
question without consulting the R help.**

```{r problem3c}
# your code here
```

d. **What is the mean height for the 1st child in each family?**

```{r problem3d}
# your code here
```

e. **Create a table showing the mean height for male and female children.**
```{r problem3e}
# your code here
```

f. **What was the average number of children each family had?**

```{r problem3f}
# your code here
```

g. **Convert the children's heights from inches to centimeters and store
it in a column called `childHeight_cm` in the `GaltonFamilies` dataset.
Show the first few rows of this dataset.**

```{r problem3g}
# your code here
```


## Problem 4: Spurious Correlation

```{r gen-data-spurious, cache = TRUE, eval=TRUE}
# set seed for reproducibility
set.seed(1234)
N <- 25
ngroups <- 100000
sim_data <- data.frame(group = rep(1:ngroups, each = N),
                       X = rnorm(N * ngroups),
                       Y = rnorm(N * ngroups))
```

In the code above, we generate `r ngroups` groups of `r N` observations
each. In each group, we have X and Y, where X and Y are independent
normally distributed data and have 0 correlation.

a. **Find the correlation between X and Y for each group, and display
the highest correlations.**

Hint: since the data is quite large and your code might take a few
moments to run, you can test your code on a subset of the data first
(e.g. you can take the first 100 groups like so):

```{r subset}
sim_data_sub <- sim_data %>% filter(group <= 100)
```

In general, this is good practice whenever you have a large dataset:
If you are writing new code and it takes a while to run on the whole
dataset, get it to work on a subset first. By running on a subset, you
can iterate faster.

However, please do run your final code on the whole dataset.

```{r cor, cache = TRUE}
# your code here
```

b. **The highest correlation is around 0.8. Can you explain why we see
such a high correlation when X and Y are supposed to be independent and
thus uncorrelated?**


# Part III: Plotting

## Problem 5

**Show a plot of the data for the group that had the highest correlation
you found in Problem 4.**

```{r problem5}
# your code here
```

Grading: 1pt.

## Problem 6

We generate some sample data below. The data is numeric, and has 3
columns: X, Y, Z.

```{r gen-data-corr}
N <- 100
Sigma <- matrix(c(1, 0.75, 0.75, 1), nrow = 2, ncol = 2) * 1.5
means <- list(c(11, 3), c(9, 5), c(7, 7), c(5, 9), c(3, 11))
dat <- lapply(means, function(mu)
  rmvnorm(N, mu, Sigma))
dat <- as.data.frame(Reduce(rbind, dat)) %>%
  mutate(Z = as.character(rep(seq_along(means), each = N)))
names(dat) <- c("X", "Y", "Z")
```

a. **Compute the overall correlation between X and Y.**

```{r problem6a}
# your code here
```

b. **Make a plot showing the relationship between X and Y. Comment on
the correlation that you see.**

```{r problem6b}
# your code here
```

Your text answer here.

c. **Compute the correlations between X and Y for each level of Z.**

```{r problem6c}
# your code here
```

d. **Make a plot showing the relationship between X and Y, but this
time, color the points using the value of Z. Comment on the result,
especially any differences between this plot and the previous plot.**

```{r problem6d}
# your code here
```

Your text answer here.


# Part IV: Bash practices

## Problem 7: Bash practices on Odyessy

Please answer the following question using bash commands and include those in 
your answer. Data are available at `/n/stats115/2020/HW1/public_MC3.maf`

Mutation Annotation Format ([MAF](https://docs.gdc.cancer.gov/Data/File_Formats/MAF_Format/)) is a tab-delimited text file with aggregated mutation information. `public_MC3.maf`
is a curated list of [somatic mutation](https://www.britannica.com/science/somatic-mutation) 
occured in many patients. Since a complete MAF file contains far mroe 
information than we need, in this problem we will focus on part of it.

```
Chromosome	Start_Position	Hugo_Symbol	Variant_Classification
10	123810032	TACC2	Missense_Mutation
10	133967449	JAKMIP3	Silent
11	124489539	PANX3	Missense_Mutation
11	47380512	SPI1	Missense_Mutation
11	89868837	NAALAD2	Missense_Mutation
11	92570936	FAT3	Silent
12	107371855	MTERFD3	Missense_Mutation
12	108012011	BTBD11	Missense_Mutation
12	117768962	NOS1	5'Flank
```

In  `/n/stats115/2020/HW1/MC3/public_MC3.maf`, `Chromosome` and `Start_Position` 
together specifies the genomics location where a location has happened. 
`Hogo_symbol` is the overlapping gene of that location, and 
`Variant_Classification` specifies how it influences downstream biological 
processes, e.g. transcription and translation. We are interested to find out
recurrent mutations with biological significance.

Hint: In this exercise you might need to use `cut` and `sort` command. Please 
try `man cut` and `man sort` to understand how it works. You might also 
benefit if you use pipes `|`.

a. How many lines are there in this file? How many times "KRAS" gene 
has appeared?

```{r engine="bash", eval=FALSE}
# your bash command
```

b. How many unique `Variant_Classification` are there in the MAF? Please 
count occurence of each type and sort them. Which one is the most frequent? 

```{r engine="bash", eval=FALSE}
# your bash command
```

c. What are the top FIVE most frequent genes? Please provide 
the bash command and equivalent Python command. If you are a PI 
looking for a gene to investigate (you need to find a gene with potentially 
better biological significance), which gene out of the top 5 would you 
choose? Why?

```{r engine="bash", eval=FALSE}
# your bash command
```

Equivalent python command:

```{r engine="python", eval=FALSE}
# your python command
```

Your text answer:

d. Write a bash program that determines whether a user-input year ([YYYY]) is 
a leap year or not (Definition: multiples of four with the exception of 
centennial years not divisible by 400). The user-input can be either positional
or interactive.

```{r engine="bash", eval=FALSE}
# your bash command
```


# Part V. High throughput sequencing read mapping

We will give you a simple example to test high throughput sequencing
alignment for RNA-seq data. Normally for paired-end sequencing data,
each sample will have two separate FASTQ files, with line-by-line
correspondence to the two reads from the same fragment. Read mapping
could take a long time, so we have created just two FASTQ files of one
RNA-seq sample with only 1M fragments (three 2 X 1M reads) for you to run STAR
instead of the full data. The files are located at
`/n/stat115/2020/HW1/loop/`. 

```
loop
├── A_l.fastq
├── A_r.fastq
├── B_l.fastq
├── B_r.fastq
├── C_l.fastq
└── C_r.fastq
```

Please include the commands that you used to run BWA and STAR in your
answers.


## Problem 8: BWA

a. Use BWA (Li & Durbin, Bioinformatics 2009) to map the reads to the
Hg38 version of the reference genome, available on Odyssey at
`/n/stat115/2020/HW1/hg38.fasta`. 

In `/n/stat115/2020/HW1/BWA/loop`, you are provided with three `.fastq` files/
Write a for loop in bash to align the paired-end reads to the reference using 
BWA on a compute node. Use the PE alignment mode and generate the output in 
SAM format. Use SAMTools on the output to find out how many fragments are 
mappable and uniquely mappable. How many rows are in each output SAM files?

```{r engine="bash", eval=FALSE}
# please copy your bash script here
```

Please provide text answer here.

b. Use slurm to submit the same BWA alignment jobs onto cluster for `A_r.fastq` 
and `A_l.fastq`. Please copy the content of your `sbatch` file here.

```{r engine="bash", eval=FALSE}
# please copy your sbatch file here
```

## Problem 9: STAR alignment

a. Use STAR (Dobin et al, Bioinformatics 2012) to map the reads to the
reference genome, available on Odyssey at
`/n/stat115/2020/HW1/STARIndex`. Use the paired-end alignment mode and
generate the output in SAM format for `A_r.fastq` 
and `A_l.fastq` in `/n/stat115/2020/HW1/loop`. STAR should have a report.  
How many fragments are mappable and how many are uniquely mappable?

```{r engine="bash", eval=FALSE}
# please copy your sbatch file here
```

b. If you are getting a different number of mappable fragments between
BWA and STAR on the same data, why?


Please provide text answer here.


# Part VII: Dynamic programming with Python

## Problem 10 Dyanic Programming

Given a list of finite integer numbers: e.g. -2, 1, 7, -4, 5, 2, -3, -6, 4, 3, 
-8, -1, 6, -7, -9, -5.
Write a python script to maximize the Z where Z is the sum of the
numbers from location X to location Y on this list. Be aware, your
algorithm should look at each number ONLY ONCE from left to right.

Hint: You can use dynamic programming to solve this problem with <20
lines of codes.

```{r engine="python", eval=FALSE}
# your bash command
```


